 ---
title: "Influences of medical data on blood glucose level"
author: "Alliance of the Unshabby"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Provide an introduction that includes:

- **Purpose of the Analysis**: Explain what you are attempting to accomplish.
- **Background Information**:
  - What is this data?
  - Where did it come from?
  - What are the variables?
  - Why is it interesting to you?
- **Objective of the Model**: Describe why you are creating a model for this data and the goal of the model.

*Note: Provide enough background so that a reader would not need to load your data to understand your report. Assume the reader is familiar with the course concepts but not your data.*

## Methods

This section contains the bulk of your work and R code. Use RMarkdown and code comments to explain your code if needed.

### Data Preparation

```{r}
file_path <- "StudentPerformanceFactors.csv"

students_data <- read.csv(file_path)

head(students_data$Exam_Score)
```

- Describe any data preparation performed on the original data before modeling.
- Include descriptions of all relevant variables.
- Explain any transformations or cleaning steps.

```{r}
# Remove empty observations
students_data = subset(students_data, students_data$Teacher_Quality != "")
students_data = subset(students_data, students_data$Parental_Education_Level != "")
students_data = subset(students_data, students_data$Distance_from_Home != "")
students_data = na.omit(students_data)

# Change Parental_Involvement to a factor variable
students_data$Parental_Involvement = as.factor(students_data$Parental_Involvement)
levels(students_data$Parental_Involvement)

# Change Access_to_Resources to a factor variable
students_data$Access_to_Resources = as.factor(students_data$Access_to_Resources)
levels(students_data$Access_to_Resources)

# Change Extracurricular_Activities to a factor variable
students_data$Extracurricular_Activities = as.factor(students_data$Extracurricular_Activities)
levels(students_data$Extracurricular_Activities)

# Change Motivation_Level to a factor variable
students_data$Motivation_Level = as.factor(students_data$Motivation_Level)
levels(students_data$Motivation_Level)

# Change Internet_Access to a factor variable
students_data$Internet_Access = as.factor(students_data$Internet_Access)
levels(students_data$Internet_Access)

# Change Family_Income to a factor variable
students_data$Family_Income = as.factor(students_data$Family_Income)
levels(students_data$Family_Income)

# Change Teacher_Quality to a factor variable
students_data$Teacher_Quality = as.factor(students_data$Teacher_Quality)
levels(students_data$Teacher_Quality)

# Change School_Type to a factor variable
students_data$School_Type = as.factor(students_data$School_Type)
levels(students_data$School_Type)

# Change Peer_Influence to a factor variable
students_data$Peer_Influence = as.factor(students_data$Peer_Influence)
levels(students_data$Peer_Influence)

# Change Learning_Disabilities to a factor variable
students_data$Learning_Disabilities = as.factor(students_data$Learning_Disabilities)
levels(students_data$Learning_Disabilities)

# Change Parental_Education_Level to a factor variable
students_data$Parental_Education_Level = as.factor(students_data$Parental_Education_Level)
levels(students_data$Parental_Education_Level)

# Change Distance_from_Home to a factor variable
students_data$Distance_from_Home = as.factor(students_data$Distance_from_Home)
levels(students_data$Distance_from_Home)

# Change Gender to a factor variable
students_data$Gender = as.factor(students_data$Gender)
levels(students_data$Gender)

# Check the size of the prepared dataset
length(students_data$Exam_Score)
```

### Modeling Process

- **Methodology**: Detail the methods you applied (e.g., multiple linear regression, dummy variables, interaction terms).
- **Decision-Making Process**:
  - Narrate your step-by-step decision-making as you adjusted the model.
  - Discuss how you attempted to validate model assumptions.
- **Diagnostics**:
  - Include residual and outlier diagnostics.
  - Discuss any transformations or model selections made.

*Your task is to use appropriate methods to find a good model that can correctly answer a question about the dataset, and then communicate your result effectively.*

We will use 10 of the predictors in the dataset to build a model. By adjusting the initial model, we will try to get a good model that can accurately predict the student's `Exam_Score` from the data of `Hours_Studied`, `Attendance`, `Parental_Involvement`, `Extracurricular_Activities`, `Sleep_Hours`, `Previous_Scores`, `Tutoring_Sessions`, `Physical_Activity`, `Parental_Education_Level`, and `Gender`.

```{r}
# Construct an initial model. We choose a additive model of 10 predictors by multiple linear regression.
initial_model = lm(Exam_Score ~ Hours_Studied + Attendance + Parental_Involvement + 
                   Extracurricular_Activities + Sleep_Hours + Previous_Scores + 
                   Tutoring_Sessions + Physical_Activity + Parental_Education_Level + Gender, 
                   data = students_data)

# Check collinearity
faraway::vif(initial_model)

# Check R^2
summary(initial_model)$r.squared
```
From the output, we can see that the variance inflation factor of all the predictors are lower than 5, so we may not worry about collinearity.

```{r warning=FALSE}
# Then, we do a AIC backward search to select a model
selected_model = step(initial_model, direction = "backward", trace = 0)

# Check the selected model
selected_model

# Check collinearity
faraway::vif(initial_model)
```
From the output of vif, we can see that we do not need to worry about collinearity for the selected model. Then, we want to check the model assumptions.

```{r}
# Construct a function for plotting
assumptions_plots = function(model, pcol = "grey", lcol = "dodgerblue") {
  par(mfrow = c(1, 2))
    plot(fitted(model), resid(model), col = pcol, pch = 20,
     xlab = "Fitted", ylab = "Residuals", main = "Fitted Versus Residuals Plot")
    abline(h = 0, col = lcol, lwd = 2)
    qqnorm(resid(model), main = "Normal Q-Q Plot", col = pcol)
    qqline(resid(model), col = lcol, lwd = 2)
}
```

```{r}
# Construct a function for calculating the LOOCV RMSE
calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}
```

```{r message=FALSE, warning=FALSE}
library(lmtest)
# Plot the Fitted vs Residual plot and Normal Q-Q plot
assumptions_plots(selected_model)

# Do a Breusch-Pagan Test
bptest(selected_model)
```

For both of the plots, we can clearly see that there are some outliers in this model, which maay be influential. Thus, we may need to do some adjustments for the selected model.From the Fitted Versus Residual Plot and the Breusch-Pagan Test, we can see that the mean of the residuals is roughly 0 and the spread of the residuals is roughly the same at most fitted values and the p-value for the Breusch-Pagan Test is really large, which means the linearity and constant variance assumptions are valid. However, from the Normal Q-Q plot, we can see that the points do not do not closely follow a straight line, which suggest that the normal distribution assumption is not valid.

```{r}
# Calculate the cooks distance of the selected model
cd_selected_model = cooks.distance(selected_model)

# Build a fixed model that excludes all the influential data (cooks distance > 4/n)
fix_model = lm(Exam_Score ~ Hours_Studied + Attendance + Parental_Involvement + 
    Extracurricular_Activities + Previous_Scores + Tutoring_Sessions + 
    Physical_Activity + Parental_Education_Level, 
    data = students_data, 
    subset = cd_selected_model <= 4 / length(cd_selected_model))
    
# Plot the Fitted vs Residual plot and Normal Q-Q plot
assumptions_plots(fix_model)
```

```{r message=FALSE, warning=FALSE}
# Do a Breusch-Pagan Test
bptest(fix_model)
```
The Normal Q-Q plot is much better and clearly shows that the majority of the points closely follow a straight line, indicating that the normal distribution assumption is correct. The Fitted Versus Residual Plot may look like that mean of the residuals is approximately 0 and the spread of the residuals is roughly the same at most fitted values. However, the Breusch-Pagan test gives a pretty low p-value: `r bptest(fix_model)$p.value`, which implies that the constant variance assumption is not valid for the model after adjustments. Therefore, we need to do some transformations.

```{r}
# Do some Predictor transformations
trans_model = lm(Exam_Score ~ Hours_Studied + I(Hours_Studied ^ 2) + log(Attendance) + Parental_Involvement + 
    Extracurricular_Activities + log(Previous_Scores) + sqrt(Tutoring_Sessions) + 
    Physical_Activity + I(Physical_Activity ^ 2) + Parental_Education_Level, 
    data = students_data, 
    subset = cd_selected_model <= 4 / length(cd_selected_model))

# Do a Breusch-Pagan Test
bptest(trans_model)

# Plot the Fitted vs Residual plot and Normal Q-Q plot
assumptions_plots(trans_model)
```

After the transformation, we can find that the p-value of the Breusch-Pagan test is `r bptest(trans_model)$p.value`, which means that the test will fail to reject for a low $\alpha$, which indicate that the constant variance assumption is valid. Moreover, from the Fitted Versus Residual Plot, we can see that the mean of the residuals is approximately 0 and the spread of the residuals is roughly the same at most fitted values, which also indicate that the linearity and constant variance assumption is valid. Moreover, the Normal Q-Q plot looks really good as the majority of the points closely follow a straight line, indicating that the normal distribution assumption is correct. Therefore, we can choose this transformed model as a result model.

## Results

- Present numerical and graphical summaries of your results.
- Report the final model you have chosen.
- Provide evidence that your final choice of model is a good one.
- Include relevant tables, charts, or figures, and discuss them in the text.

## Discussion

- Discuss your results in the context of the data.
- Explain how your final model is useful.
- Reflect on the implications of your findings.

## Appendix

Include supplementary materials that support your report but are not essential to the main text.

- Additional code and analysis that would clutter the main report.
- Extra figures or tables not included in the Results section.
- Any other relevant information.

*Do not simply dump code here; only include supplementary material.*

### Group Members

- Anshul Kaushik
- Desmond Fung
- Jon Green
- Kangyi Jiang

