---
title: "STAT 420 Fall 2024 Project Proposal"
author: "Alliance of the Unshabby"
date: "2024-12-10"
output: html_document
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

## 1. Group Members

- Anshul Kaushik
- Desmond Fung
- Jon Green
- Kangyi Jiang

## 2. Tentative Title

### Influences of student data on final exam scores

## 3. Dataset Description

We have selected a dataset that provides various factors potentially influencing students’ final exam performance. This dataset includes demographic information, academic habits, environmental influences, and family background details for each student. The data is substantial, with around 20 variables and 6607 records. Key variables of interest include:

- **Exam_Score:** The response variable measuring final exam performance.
- **Hours_Studied:** Numeric measure of the average daily hours spent studying.
- **Attendance:** Numeric measure indicating the number of lectures or classes attended.
- **Parental_Involvement:** Categorical factor (High, Medium, Low) reflecting the level of parental support.
- **Access_to_Resources:** Categorical factor indicating the availability of learning materials (High, Medium, Low).
- **Extracurricular_Activities:** Categorical factor (Yes/No) representing participation in non-academic school activities.
- **Previous_Scores:** Numeric measure of prior academic performance.
- **Motivation_Level:** Categorical factor (High, Medium, Low) reflecting the student’s self-reported drive.
- **Parental_Education_Level:** Categorical factor indicating the highest education level achieved by the parents (High_School, Undergraduate, Postgraduate).
- **Distance_from_Home:** Categorical factor (Close, Moderate, Far) measuring proximity between home and school.
  
These are only some of the variables from a rich dataset designed for educational analytics. We will focus primarily on understanding how select predictors—such as study habits, parental involvement, and resource availability—affect final exam scores.

## 4. Background Information

The dataset we have chosen is the "Student Performance Factors" dataset, which is publicly available on Kaggle. It is a synthetic dataset created to simulate realistic student performance scenarios, making it ideal for exploratory analysis and modeling approaches in an academic context.

**Data Source:** [Student Performance Factors on Kaggle](https://www.kaggle.com/datasets/lainguyn123/student-performance-factors)

This dataset aggregates hypothetical student records that include a combination of academic, personal, and environmental factors. Although the data is not sourced from real-world institutions, it has been designed to provide plausible distributions and relationships. The use of this synthetic dataset allows for free exploration without confidentiality concerns and serves as an excellent resource for examining what might influence exam outcomes.

## 5. Statement of Interest

Our goal is to explore how various factors—ranging from hours studied and parental involvement to resource accessibility and extracurricular engagement—contribute to predicting final exam scores. By modeling these relationships, we hope to better understand which factors hold the strongest association with academic success. This insight can guide targeted interventions, help educators and policymakers prioritize resources, and offer students evidence-based strategies to improve their learning experience. Ultimately, we aim to produce a model that can shed light on key predictors of student performance and facilitate more informed decisions in educational settings.

## 6. Evidence of Data Loading into R

Below is a short snippet of R code to load and quickly inspect the data. We will use this code to confirm that we can read the dataset and access the `Exam_Score` variable.

```{r}
file_path <- "StudentPerformanceFactors.csv"

students_data <- read.csv(file_path)

head(students_data$Exam_Score)
```

This code should output: "[1] 67 61 74 71 70 71". 
