---
title: "STAT 420 Fall 2024 Project Proposal"
author: "Alliance of the Unshabby"
date: "2024-11-16"
output: html_document
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```


## 1. Group Members

- Anshul Kaushik
- Desmond Fung
- Jon Green
- Kangyi Jiang

## 2. Tentative Project Title

### Influences of medical data on blood glucose level

## 3. Dataset Description

We chose a dataset that includes patients' medical and demographic information, as well as their diabetes status (positive or negative). The dataset is huge, with 9 variables and 100,000 observations. The data includes variables such as age, gender, body mass index (BMI), hypertension, heart disease, smoking history, HbA1c level, and blood glucose level.

The description of variables:

1. gender: Gender refers to a person's biological sex, which may influence their vulnerability to diabetes.

2. age: Age is an important factor as diabetes is more commonly diagnosed in older adults. Age ranges from 0-80 in the dataset.

3. hypertension: Hypertension is a medical condition in which blood pressure in the arteries is consistently high. It has two possible values: 0 and 1, with 0 indicating that they do not have hypertension and 1 indicating that they have.

4. heart_disease: Heart disease is another medical condition linked to an increased chance of acquiring diabetes. It has a value of either 0 or 1, with 0 denoting no heart disease and 1 indicating heart disease.

5. smoking_history: Smoking history is also considered a risk factor for diabetes and can exacerbate the complications associated with diabetes. There are 5 categories in the dataset i.e not current,former,No Info,current,never and ever.

6. bmi: BMI (Body Mass Index) is a calculation of body fat based on weight and height. Increased BMI values are associated with an increased risk of diabetes. The dataset includes BMI values ranging from 10.16 to 71.55. Underweight is defined as a BMI of less than 18.5, normal as 18.5-24.9, overweight as 25-29.9, and obese as 30 or higher.

7. HbA1c_level: HbA1c (Hemoglobin A1c) is a measurement of a person's average blood sugar level over the previous 2-3 months. Higher levels imply an increased risk of getting diabetes. Diabetes is typically diagnosed when HbA1c levels exceed 6.5%.

8. blood_glucose_level: Blood glucose level is the amount of glucose in the bloodstream at any particular time. High blood glucose levels are a key sign of diabetes.

9. diabetes: Diabetes is a predicted variable in the dataset, with 1 indicating the occurrence of diabetes and 0 indicating its absence.

## 4. Background Information

The Diabetes Prediction Dataset is publicly available on Kaggle, contributed by iammustafatz ([Kaggle Source](https://www.kaggle.com/datasets/iammustafatz/diabetes-prediction-dataset)). It contains medical and demographic data, including features like age, gender, BMI, blood glucose levels, HbA1c levels, and smoking history, which are crucial for predicting diabetes risk.

The dataset is primarily sourced from Electronic Health Records (EHRs) aggregated from multiple healthcare providers. EHRs provide comprehensive patient data, enabling the identification of patterns and trends in health status. Data was collected via surveys, medical records, and laboratory tests, then cleaned and processed for consistency and quality.

Due to confidentiality restrictions, specific details about the data's origins cannot be disclosed. However, this dataset has been widely used in research and serves as a valuable resource for building predictive models and exploring diabetes-related insights.

## 5. Statement of Interest

We’re exploring the **Diabetes Prediction Dataset** to better understand what factors influence **blood glucose levels**. 

By focusing on key variables like `age`, `BMI`, `hypertension`, `heart_disease`, `smoking_history`, and `HbA1c_level`, we’ll use linear regression to analyze how these factors contribute to predicting blood glucose levels.

Our goal is simple: uncover patterns that can help identify early warning signs for diabetes-related issues. This kind of insight could make a real difference in creating **personalized care strategies** and improving how we approach **diabetes prevention and management**.


## 6. Evidence of Data Loading in R

```{r}
file_path <- "diabetes_prediction_dataset.csv"

diabetes_data <- read.csv(file_path)

head(diabetes_data$blood_glucose_level)
```

This code should output: "[1] 140  80 158 155 155  85". 

