---
title: "Influences of medical data on blood glucose level"
author: "Alliance of the Unshabby"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Provide an introduction that includes:

- **Purpose of the Analysis**: Explain what you are attempting to accomplish.
- **Background Information**:
  - What is this data?
  - Where did it come from?
  - What are the variables?
  - Why is it interesting to you?
- **Objective of the Model**: Describe why you are creating a model for this data and the goal of the model.

*Note: Provide enough background so that a reader would not need to load your data to understand your report. Assume the reader is familiar with the course concepts but not your data.*

## Methods

This section contains the bulk of your work and R code. Use RMarkdown and code comments to explain your code if needed.

### Data Preparation

```{r}
file_path <- "diabetes_prediction_dataset.csv"

diabetes_data <- read.csv(file_path)

head(diabetes_data$blood_glucose_level)
```

- Describe any data preparation performed on the original data before modeling.
- Include descriptions of all relevant variables.
- Explain any transformations or cleaning steps.

### Modeling Process

- **Methodology**: Detail the methods you applied (e.g., multiple linear regression, dummy variables, interaction terms).
- **Decision-Making Process**:
  - Narrate your step-by-step decision-making as you adjusted the model.
  - Discuss how you attempted to validate model assumptions.
- **Diagnostics**:
  - Include residual and outlier diagnostics.
  - Discuss any transformations or model selections made.

*Your task is to use appropriate methods to find a good model that can correctly answer a question about the dataset, and then communicate your result effectively.*

## Results

- Present numerical and graphical summaries of your results.
- Report the final model you have chosen.
- Provide evidence that your final choice of model is a good one.
- Include relevant tables, charts, or figures, and discuss them in the text.

## Discussion

- Discuss your results in the context of the data.
- Explain how your final model is useful.
- Reflect on the implications of your findings.

## Appendix

Include supplementary materials that support your report but are not essential to the main text.

- Additional code and analysis that would clutter the main report.
- Extra figures or tables not included in the Results section.
- Any other relevant information.

*Do not simply dump code here; only include supplementary material.*

### Group Members

- Anshul Kaushik
- Desmond Fung
- Jon Green
- Kangyi Jiang

