---
title: "STAT 420 Fall 2024 Project Proposal"
author: "Alliance of the Unshabby"
date: "2024-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Group Members

- Anshul Kaushik
- Desmond Fung
- Jon Green
- Kangyi Jiang

## 2. Tentative Project Title

## 3. Dataset Description

Description of the data file (what they contain including number of variables and number of records). You do not necessarily have to list all the variables, but at least mention those of greatest importance.

## 4. Background Information

The Diabetes Prediction Dataset is publicly available on Kaggle, contributed by iammustafatz ([Kaggle Source](https://www.kaggle.com/datasets/iammustafatz/diabetes-prediction-dataset)). It contains medical and demographic data, including features like age, gender, BMI, blood glucose levels, HbA1c levels, and smoking history, which are crucial for predicting diabetes risk.

The dataset is primarily sourced from Electronic Health Records (EHRs) aggregated from multiple healthcare providers. EHRs provide comprehensive patient data, enabling the identification of patterns and trends in health status. Data was collected via surveys, medical records, and laboratory tests, then cleaned and processed for consistency and quality.

Due to confidentiality restrictions, specific details about the data's origins cannot be disclosed. However, this dataset has been widely used in research and serves as a valuable resource for building predictive models and exploring diabetes-related insights.

## 5. Statement of Interest

A brief statement of the business, science, research, or personal interest you have in the data set which you hope to explore

## 6. Evidence of Data Loading in R

```{r}

```

